Optogenetic Maxwell Demon to Exploit Intrinsic Noise and Control Cell Differentiation Despite Time Delays and Extrinsic Variability

### 2023-06-23 paper.tex
updated figs
before making major text edits'
update

Unnderstanding the effects of noise in control theory is important to ()
Much work has been done to make control performance robust to noise, but the development of controllers which use noise to their advantage is not fully understood.
Motivated by optogentics, we proposed a difficult control problem which requires the exploitation of noise in a stochastic system to break symmetry between two signals.
in previous work, we found multiple such controllers which could exploit noise to break symmetry between two cells under a variety of system information.
This work extends that analysis to include stochastic systems that trace a moving target, are affected by time delay, intrinsic noise, or measurement uncertainty.

@@

\section{Introduction}
\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{CartoonAndControler.pdf}
\caption{{\bf Optogenetic control of multiple cells using a single light input.}
({\bf A}) Schematic of the light activated genetic system with autoregulation.
({\bf B}) Diagram of the stochastic SIMO control problem using two optogenetic cells sharing an input.
({\bf C}) Adapted from (PAPER). Noise exploiting controllers were optimized to a fully aware control input (I), and partially aware control input(III). Marginal distributions are solved using the finite state projection and show a break in symmetry using both controllers (II and IV) }
\label{cartoons}
\end{center}
\vspace{-0.3in}
\end{figure}
% XXX - Please modify figure as follows: (1) switch panels A and B to match caption. (2) Change title of panel B (previously panel A) to "External Observer, controller and actuator".  (3) Change title of panel A (previously panel B) to "Internal gene regulation pathway". (4) Change title of panel C to "Feedback control law and resulting performance"
% MMM - Made the edit

Synthetic biology seeks to create modular\cite{Ng2019} and orthogonal\cite{Liu2018} components to sense and actuate\cite{Sheets2020} complex logical systems\cite{Groseclose2020}, and which are capable of performing a variety of advanced biological behaviors\cite{Shin2020}. 
New optogenetic tools have increased our ability to actuate embedded systems within cells reliably and with strong control performance\cite{Sheets2020,Baumschlager2017,Chen2020,Lillacci2018}. Advances in these two fields have enabled computer programmable control of cellular protein production through the use of external optogenetic inputs and smart microscope techniques\cite{Fox2021,Baumschlager2021,Lugagne2017}. These types of digital-synthetic actuators allow for fine-tuned, computer-modulated control of cellular systems that were previously impossible\cite{Rullan2018, Baumschlager2017} and with fast response times in comparison to chemical diffusion.

Classical and modern control methods like PID control and model predictive control have been implemented in such systems\cite{} which can control synthetic systems to different stable points. 
% XXX Missing Citation above.
Theoretical works have shown that stochastic systems can be controlled to unstable points by oscillating around those points \cite{Guarino2020} but new control techniques which exploit the full probability distribution information of the system can be used in certain scenarios to control simple systems to unstable points by exploiting the natural noise of protein production to its advantage. Previous control effort focus on deterministic ODE formulations of the control problem, but only a few studies seek to analyze the full chemical master equation to exploit noise to {\em improve} the control of synthetic biology processes \cite{Szymanska2015,May2021}.

Szymanska et al.~\cite{Szymanska2015} examined a theoretical system inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, where two gene products repress one another, but where degradation of one protein was under the control of UV radiation via the XXX pathway. 
% XXX - add citation to Kobayashi paper, and update name of the pathway.  I think it was the SOS pathway.
Their analysis showed that control was possible despite certain parameter errors and time delays due to maturation of fluorescent proteins and limited observation of the regulatory proteins. 
To explore this concept in the context of more realistic models and parameters, May et al.~\cite{May2021} fit a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure\ \ref{cartoons}A, top promoter). 
% XXX - add details about the promoter and citation.
% XXX - let's change the order of panels A and B for this figure so that we can refer to them in the proper order.  I adjusted the text below to assume this change.
An extended model showed that addition of a positive auto-regulation module (Figure\ \ref{cartoons}A, bottom promoter) could help to maintain an elevated expression phenotype in the presence of UV excitation.
% XXX - Missing figure reference
Using discrete stochastic models based on direct solutions to the chemical master equation, it was shown that the combination of biochemical noise, nonlinear auto-regulation, and a {\em single} optogenetic feedback could allow two genetically identical cells with arbitrary initial conditions to be controlled simultaneously to reach different specified phenotypes (Figure \ref{cartoons}B). 
For example, Figure\ \ref{cartoons}C(left) reproduces the heat maps for two control laws, where the level of a single external UV excitation signal was modulated depending on the observed cellular activity, and Figure\ \ref{cartoons}C(right) shows the corresponding marginal stationary distributions that could be achieved under these control laws. 
Effective control was possible not only for a Fully Aware Feedback Controller (FAC, Figure\ \ref{cartoons}C(top)) that observed the protein counts of all observed cells, but control could be achieved nearly as well for a Partially Aware Feedback Controller (PAC, Figure\ \ref{cartoons}C(bottom)) that observed the protein counts of only one cell. 
Moreover, it was again seen that these control laws remained effective even in the circumstance where the model used to design the control law was an inexact simplification of the process dynamics to which the controller was later applied.  
%The solutions of the controllers was carried out by solving the chemical master equation (i.e., the forward Komogorov equation) to estimate the probability of possible cellular states. For a two cell, single-input-single output (SIMO) control problem, the FAC performed the best (figure\ \ref{cartoons}C(top)), and the PAC (figure\ \ref{cartoons}C(bottom)) 2nd, and pMPC controllers 3rd. In all cases, the controller's knowledge and explicit treatment of intrinsic noise was critical to improve the control performance, break symmetry, and enable the SIMO control of all cells.

In this paper, the analysis of this Single-Input-Multiple-Output (SIMO) multicellular control problem is extended to examine how the FAC and PAC controllers are affected by additional model uncertainties, including coarse grained approximations to the system dynamics, uncertainties or errors in system parameters, and delays in the biochemical sensors or optogenetic controllers through which the external feedback is implemented. In the next section, we briefly describe our methods to formulate the chemical master equation to analyze multiple individual cells in the same spatially homogeneous environment; we define the FAC and PAC controllers and their effects the dynamics of cells within that environment; and we explain how the control law is optimized to achieve desired performance objectives. Next, in the `Results' section, we explore how granularity of the model formulation, inaccuracy of parametric assumptions, or introduction of time delays would affect the FAC and PAC control performance. Finally, in `Conclusions' we summarize our findings and discuss the implications that these controllers, and their robustness to modeling errors, may have on future systems in synthetic biology.

\section{Methods}
In previous work~\cite{May2021}, a six-species gene regulation model was developed to describe light-activated association of two T7 split domains (species 1 and 2) to form active T7 polymerases (species 3) which could then in turn associate with inactive T7 promoters (species 4) to create an active gene construct (species 5) that would produce its eventual protein product (species 6). A second, far simpler single-species model was also developed by assuming quasi-steady equilibrium for the first five species. Both models were independently parametrized by fitting to the same experimental data from Baumschlager, et al.~\cite{XXX}, and then extended to include a secondary self-activated promoter-gene construct for which the expression rate was defined by a Hill function (see Eqn.\ \ref{prodRate} below). In \cite{May2021}, we showed that a control law identified using the simple model would work to achieve differential control for the more complex model, thus motivating our current study's efforts to explore further how control performance would change upon additional simplifying assumptions or errors in model specification.  In the following subsections, we discuss our methods to formulate this model, show how it is analyzed numerically, and discuss the specification of the control objective and the determination of an optimal controller.

\subsection{Model}
To examine how model approximations would affect the application of noise-enhanced control strategies, we start with the one species model from \cite{May2021} (see Fig.\ \ref{cartoons}A), which consists of two reactions: protein production and degradation.  The nonlinear light-activated protein production rate is given by
 \begin{equation}
\nu_1(x,t)= \kappa\frac{x^\eta}{x^\eta+\beta^\eta}+ k_0 + u(UV(t)),
\label{prodRate}
\end{equation}
%
%These model dynamics were then simplified to a single ODE with auto-regulation describing the protein accumulation as: 
%\begin{equation}
%\frac{dp}{dt}=u(x_1,x_2)  + \kappa\frac{p^\eta}{p^\eta+\beta^\eta}+ k_0 -\gamma p,
%\label{rateeq}
%\end{equation}
%
where $x$ is the instantaneous protein level; $\kappa$ is the maximum strength of the auto-regulation promoter; $\beta$ is the concentration at which auto-regulation promoter reaches half its strength; $\eta$ is the cooperativity in the auto-regulation promoter; $k_0$ is the promoter leakage rate; and $u(UV(t))$ is the T7 promoter strength. Through external feedback, the level of light excitation, and therefore the T7 promoter strength, $u(UV(t))$, can be modulated in response to the state of the system, thus removing explicit dependence on time. For example, in a system with the observation of two cells, one could define $u(UV(t)) = u^{\rm (FAC)}(x_1,x_2)$ as shown in Fig.\ \ref{cartoons}C(top left) for the fully aware control law, and for a system with the observation of only one cell, one could define $u^{\rm (PAC)}(UV(t)) = u(x_1)$ as shown in Fig.\ \ref{cartoons}C(bottom left) for the partially aware control law. Protein degradation is assumed to be a first order process with rate $\gamma$:
\begin{equation}
\nu_2(x) = \gamma x.
\end{equation}
Parameters describing the auto-regulation promoter, $\kappa$, $\eta$, $\beta$, and $k_0$, and the degradation rate $\gamma$
%were chosen to reproduce dynamics measured in \cite{Baumschlager2017} and 
% XXX - I don't think that this is right.  Only the parameters u(UV) and gamma were fit to data.
% MMM - yes that is right
are presented in Table 1~\cite{May2021}.

@@

With these stoichiometry and propensities, we can formulate and run the Gillespie stochastic simulation algorithm\cite{Gillespie1992,Gillespie1977} to generate representative trajectories of the stochastic process. However, quantifying performance and designing an optimal controller requires a more direct analysis of the chemical master equation (CME).  For the same specifications of the stoichiometry and propensity functions, the CME is a linear ODE that describes how probability mass changes in time due to the specified reaction propensities and stoichiometries, which can be written:
\begin{equation}
\frac{d}{dt}P(\mathbf{X}_i)=\sum_{\mu=1}^{2{\rm N_c}}
\left(
-w_{\mu}(\mathbf{X}_i)P(\mathbf{X}_i)
+w_{\mu}(\mathbf{X}_i-\mathbf{s}_{\mu})P(\mathbf{X}_i-\mathbf{s}_{\mu})
\right).\label{CMEindex}
\end{equation}
For convenience, the CME can also be formulated more compactly in matrix/vector form as:
\begin{equation}
\frac{d}{dt}\mathbf{P}=(\mathbf{A}_0+\textbf{Bu}^{\mathcal{C}})\mathbf{P},\label{CME}
\end{equation}
where $\mathbf{P} = [P(\mathbf{X}_1), P(\mathbf{X}_2), \ldots ]^T$ is the enumerated probability mass vector for all possible states of the system; $\mathbf{A}_0$ is the infinitesimal generator of the stochastic process due to the autoregulation promoter and degradation events;  $\textbf{u}^{\mathcal{C}} =[u^{\mathcal{C}}(\mathbf{X}_1), u^{\mathcal{C}}(\mathbf{X}_2), \ldots ]^T$ is the collection of control inputs associated with each state; and $\textbf{Bu}^{\mathcal{C}}$ is the contribution that these control inputs make to the infinitesimal generator when included into the feedback process.

@@

For the single-input-multiple-output control of the multiple cell system, any input signal that is applied to one cell in the population will be felt by all cells in the population. As a result, it is not possible to independently control any individual cell without introducing changes to the behavior of the others. Therefore, an effective controller must be able to strike a balance among the desired behaviors of all cells in the system.

@@

In this work, we explore two different controller designs that allow for the controller to use two different types of information. A fully aware controller ($\mathbf{u}^{FAC}$) uses direct observation of the protein count of two cells simultaneously while making its control input decision, while the partially aware controller $\mathbf{u}^{PAC}$) uses only the direct observation of a single cell:
\begin{equation}
\mathbf{u}^{FAC}=u(x_1,x_2)
\end{equation}
\begin{equation}
\mathbf{u}^{PAC}=u(x_1)
\end{equation}
% XXX - Do we need to use 0 and 1.  Why not use 1 and 2 so that this can match better to the formulaton above?  I changed them to x1 and x2 here, but please chekc to make sure that this is sonsistent throughout text and figures.
% MMM - Made the change to x1 x2
where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the number of proteins in cell one and cell two. In \cite{May2021}, the controllers $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ controllers were optimized to minimize $J$, and both $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ were saved as lookup tables after being optimized. In what follows, we will use these controllers exactly and without modification to explore how performance changes when the controllers are applied to systems that are approximations of the model under they were originally designed.

\subsection{Scaling for system granularity}
One of the major computational obstacles in using the CME or FSP formulations to analyze discrete stochastic chemical kinetics is that the dimension of the FSP scales linearly with each chemical species, and therefore exponentially with the number of independent species. To circumvent this issue, several previous studies have projected the FSP onto lower dimensional spaces based using various techniques such as time scale separations\cite{Peles2006}, Krylov subspaces\cite{Macnemara&Sidje?}, coarse meshes\cite{Munsky2008IEEE,Tapia2012CDC}, or principle orthogonal decompositions\cite{Vo2019}.
% XXX - would be good to look for a few other examples outside of our own papers where the CME dimension is reduced by linear projections.
On the other hand, an experimental obstacle to using CME analyses is that gene expression quantification based on fluorescence intensities (e.g., using fluorescent protein reporters) usually does not provide molecular resolution, but rather measures only relative changes in expression over time.

To address how variations in model or observation scales might impact the effectiveness of noise-based control, we use the controller optimized at one resolution and ask how effective it would be when applied to systems with the same dynamics but with different resolution. 
To accomplish this rescaling, let $M$ denote some metric for the size of the original system (e.g., the average number of proteins at steady state and maximal UV) and let $M'$ denote the same metric for the system at a different resolution.  
Toward this goal, we define a granularity parameter ($\alpha=M'/M$) that linearly scales the species populations to increase ($\alpha>1$) or decrease ($\alpha<1$) the maximum populations size, while maintaining the dynamics and general behavior of the model. To apply the granularity parameter, we assume that each propensity function, $w_{\mu}$, from Eqns.\ \ref{Props} is rescaled to a different level of discreteness by substituting
\begin{equation}
w'(\mathbf{X})=w(\mathbf{X}/\alpha).
\end{equation}
For example, the production and degradation of protein in cell one would become:
\begin{equation}
w'_1(\mathbf{X}) = u(\mathbf{X}/\alpha,t)  + \kappa \frac{(x_1/\alpha)^\eta}{(x_1/\alpha)^{\eta}+\beta^{\eta}} + k_0,\ w'_2(\mathbf{X}) = \gamma x_1/\alpha.
\end{equation}
% XXX - Please check this against what you did computationally. 
After this rescaling, if $\alpha<1 $ the system dynamics will now occur over a smaller range of protein counts, and if $\alpha >1$ the system dynamics occurs over a larger range of protein counts.

We note that in transforming the propensity functions, the inputs to the controller have also be scaled by $1/\alpha$ before computing the appropriate functions of that state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$. Because identification of the original control formulation, $\mathbf{u}^{FAC}(x_1,x_2)$ and $\mathbf{u}^{PAC}(x_1)$ only considered integer values for $(x_1,x_2)$, fractional number inputs for after rescaling $(x_1/\alpha,x_2/\alpha)$ are handled by 2D cubic interpolation of the nearest values.
%\begin{equation}
%u'=u(x_0 / \alpha, x_1 / \alpha) 
%\end{equation}
%\begin{equation}
%u'=u(\alpha x_0)
%\end{equation}
Finally, to provide a consistency relative scoring, the definition of the performance score is also adjusted according to sale magnitudes, i.e.,:
{\begin{align}
J' &= \sum_{i=0}^{M'}  \sum_{j=0}^{M'}P^*(x_1=i,x_2=j) ((i/\alpha - \mathbf T_1)^2 + (j/\alpha -\mathbf T_2)^2),\nonumber \\
& =\mathbf{C}'\mathbf{P}^*
\label{EuclidV}
\end{align}}

\subsection{Observation and actuation time delays}
No real optogenetic control system can provide instantaneous feedback -- there will always be delays in biochemical reactions needed for observations (e.g., transcription and translation dynamics as well as folding and maturation of fluorescent proteins), delays in data analysis and decision making (e.g., image processing and calculation of the corresponding control signals), and delays in the actuation dynamics (e.g., activation of optogenetically excitable molecules).  To explore the impacts of delays between the time a cellular fluctuation occurs and the time at which the corresponding control action can take effect, we developed a simple time-delay stochastic simulation algorithm. In this analysis, the state history is recorded following every stochastic event, allowing for reconstruction of the piecewise constant history of the species' populations, e.g., $x_1(s)$ and $x_2(s)$ for all $s\in [0,t]$.  Using this information, the time-delayed control signal at time $t$ can be specified as: 
\begin{equation}
u_{\tau}(t)=\left\{
\begin{array}{rl}
      0 ,&\text{ for }  t \leq \tau, \\
      {u}^{\mathcal{C}}(x_1(t-\tau), x_2(t-\tau)) , &\text{ for }   t > \tau,\\
\end{array}\right. 
\label{timeDelaySSA}
\end{equation}
where $\tau$ is the time delay between observation and actuation, and ${u}^{\mathcal{C}}$ is the previously optimized control law (e.g., $\mathbf{u}^{\rm FAC}$ or $\mathbf{u}^{\rm PAC}$). We note that the time delay stochastic process was only simulated using the SSA because to our knowledge an appropriate direct FSP/CME integration procedure has not yet been developed.

@@

@@

\subsection{Effects of Parameter Errors or Extrinsic Uncertainties}
\begin{figure*}
\begin{center}
\includegraphics[width=1\textwidth]{ParameterPerturbation.pdf}
\caption{Parameter sweeps using the FAC and PAC show a broad range of control performance in cell 1 ({\bf A}),  in cell 2 ({\bf B}), and in both cells ({\bf C}). }
\label{Parameter}
\end{center}
\end{figure*}

>>>>

Unnderstanding the effects of noise in control theory is important to ()
Much work has been done to make control performance robust to noise, but the development of controllers which use noise to their advantage is not fully understood.
Motivated by optogentics, we previously proposed a difficult control problem which requires the exploitation of noise in a stochastic system to break symmetry between two signals. We found multiple such controllers which could exploit noise to break symmetry between two cells under a variety of system information.
This work extends that analysis to include stochastic systems that trace a moving target, are affected by time delay, intrinsic noise, or measurement uncertainty.

@@

\section{Introduction}
\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{Cartoons.pdf}
\caption{{\bf Optogenetic control of multiple cells using a single light input.}
({\bf A}) Schematic of the light activated genetic system with autoregulation.
({\bf B}) Diagram of the stochastic SIMO control problem using two optogenetic cells sharing an input.
({\bf C}) Adapted from (PAPER). Noise exploiting controllers were optimized to a fully aware control input (I), and partially aware control input(III). Marginal distributions are solved using the finite state projection and show a break in symmetry using both controllers (II and IV) }
\label{cartoons}
\end{center}
\vspace{-0.3in}
\end{figure}
% XXX - Please modify figure as follows: (1) switch panels A and B to match caption. (2) Change title of panel B (previously panel A) to "External Observer, controller and actuator".  (3) Change title of panel A (previously panel B) to "Internal gene regulation pathway". (4) Change title of panel C to "Feedback control law and resulting performance"
% MMM - Made the edit

Noise is a common theme throughout many areas of science, including control theory XXX, gene regulation XXX, mRNA dynamics XXX, and synthetic biology XXX. Excellent work has been done to enhance the robustness of control processes against noise. However, these systems are often controlled in spite of noise rather than working with it. Many stochastic systems can be modeled using a noise-free approach with additive noise. Nevertheless, a comprehensive analysis using the full master equation can uncover interesting properties, potentially leading to novel control mechanisms.

The cell's drive towards homeostasis involves dealing with the inherently chaotic and noisy processes that reside within it. Brownian motion, stochastic gene regulation, and mRNA dynamics collectively generate a fundamentally noisy environment that cells must effectively manage. Despite these challenges, cells generally demonstrate strong capability to control these noisy processes. However, stochastic control theory is not well-developed, despite having substantial knowledge on how to model them.

Synthetic biology aims to develop modular\cite{Ng2019} and orthogonal \cite{Liu2018} components to sense and manipulate \cite{Sheets2020} complex logical systems, enabling them to exhibit a wide range of advanced biological behaviors\cite{Shin2020}. Advancements in optogenetics have enhanced our ability to reliably actuate embedded systems within cells, offering precise control\cite{Sheets2020,Baumschlager2017,Chen2020,Lillacci2018}. These developments have facilitated computer-programmable regulation of cellular protein production through external optogenetic inputs and smart microscope techniques\cite{Fox2021,Baumschlager2021,Lugagne2017}. These digital-synthetic actuators enable fine-tuned, computer-modulated control of cellular systems, previously unattainable, with faster response times compared to chemical diffusion\cite{Rullan2018, Baumschlager2017} . Classical and modern control methods like PID control and model predictive control have been implemented in such systems\cite{} which can control synthetic systems to different stable points. 
% XXX Missing Citation above.
Theoretical works have shown that stochastic systems can be guided towards unstable points by oscillating around those points \cite{Guarino2020}. However, new control techniques that leverage the complete probability distribution information of the system can exploit the inherent noise of protein production to achieve better control. Previous control efforts primarily focused on deterministic ordinary differential equation (ODE) formulations of the control problem, only a limited number of studies have delved into analyzing the full chemical master equation to harness noise to improve the control of biological processes \cite{Szymanska2015,May2021}.

inspired by the genetic toggle switch from Kobayashi et al.\cite{Kobayashi:2004}, Szymanska et al.~\cite{Szymanska2015} showed that stochastic control was possible despite certain parameter errors and time delays due to maturation of fluorescent proteins and limited observation of the regulatory proteins. 
To explore this concept in the context of more realistic models and parameters, May et al.~\cite{May2021} fit a simplified stochastic model to reproduce data measured in Baumschlager, et al.\cite{} for the expression from the XXX promoter under the optogenetic control of a UV-activated T7 polymerase (see model in Figure\ \ref{cartoons}A, top promoter).  An extended model showed that addition of a positive auto-regulation module (Figure\ \ref{cartoons}A, bottom promoter) could help to maintain an elevated expression phenotype in the presence of UV excitation.

Using discrete stochastic models based on the chemical master equation, it was demonstrated that combining biochemical noise, nonlinear auto-regulation, and a single optogenetic feedback could control two genetically identical cells with different initial conditions to reach different desired signals at different frequency and phase (Figure \ref{cartoons}B). Figure\ \ref{cartoons}C(left) presents heat maps for two control laws, which a single external UV excitation signal based on observed cellular state.  Figure\ \ref{cartoons}C(right) shows the corresponding achievable marginal stationary distributions. Effective control was possible with both Fully Aware Feedback Controllers (FAC, Figure\ \ref{cartoons}C(top)) and Partially Aware Feedback Controllers (PAC, Figure\ \ref{cartoons}C(bottom)), even when the control model was an approximate simplification of the actual process dynamics.

Using discrete stochastic models based on the chemical master equation, this study demonstrates simultaneous control of two genetically identical cells with different initial conditions, by exploiting noise, nonlinear auto-regulation, and using a single input.  Heat maps in Figure \ref{cartoons}C(left) illustrate two control laws based on observed cellular activity, while Figure\ \ref{cartoons}C(right) shows the corresponding achievable marginal stationary distributions. Effective control was achieved using both Fully Aware Feedback Controllers (FAC, Figure\ \ref{cartoons}C(top)) and Partially Aware Feedback Controllers (PAC, Figure\ \ref{cartoons}C(bottom)).

%The solutions of the controllers was carried out by solving the chemical master equation (i.e., the forward Komogorov equation) to estimate the probability of possible cellular states. For a two cell, single-input-single output (SIMO) control problem, the FAC performed the best (figure\ \ref{cartoons}C(top)), and the PAC (figure\ \ref{cartoons}C(bottom)) 2nd, and pMPC controllers 3rd. In all cases, the controller's knowledge and explicit treatment of intrinsic noise was critical to improve the control performance, break symmetry, and enable the SIMO control of all cells.

This paper extends the analysis of the Single-Input-Multiple-Output (SIMO) multicellular control problem by examining the impact of model uncertainties on FAC and PAC controllers. These uncertainties include coarse-grained system dynamics, parameter errors, and controller delays. Our methods involve formulating the chemical master equation to analyze cells in a state space, defining the controllers and their effects on cell dynamics, and optimizing the control input for desired performance. In the 'Results' section, we explore the effects of model granularity Figure \ref{cartoons}D parameter inaccuracies, and time delays Figure \ref{cartoons}E on FAC and PAC performance and add a simple scheme for controlling cells to dynamically changing point.

\section{Methods}
A gene regulation model consisting of six species was developed by May et al. \cite{May2021}. It described the light-activated association of two T7 split domains (species 1 and 2) to form an active T7 polymerase (species 3). These active polymerase could then associate with inactive T7 promoters (species 4), resulting in the formation of an active gene (species 5) that produced the desired protein product (species 6).

Additionally, a simple single-species model was developed, assuming quasi-steady equilibrium for the first five species. Both models were independently parameterized using experimental data from Baumschlager et al. \cite{XXX}. Furthermore, an extension was made to incorporate a secondary self-activated promoter-gene construct, where the expression rate was determined by a Hill function \ref{prodRate}.

\subsection{Model}
To assess the impact of model approximations on the implementation of noise-enhanced control strategies, we begin with the one-species model proposed by May et al. \cite{May2021} (refer to Fig. \ref{cartoons}A). This model comprises of two reactions: nonlinear protein production and degradation and it is defined by the following equation:
 \begin{equation}
\nu_1(x,t)= \kappa\frac{x^\eta}{x^\eta+\beta^\eta}+ k_0 + u(UV(t)),
\label{prodRate}
\end{equation}
%
%These model dynamics were then simplified to a single ODE with auto-regulation describing the protein accumulation as: 
%\begin{equation}
%\frac{dp}{dt}=u(x_1,x_2)  + \kappa\frac{p^\eta}{p^\eta+\beta^\eta}+ k_0 -\gamma p,
%\label{rateeq}
%\end{equation}
%
where $x$ is the instantaneous protein level; $\kappa$ is the maximum strength of the auto-regulation promoter; $\beta$ is the concentration at which auto-regulation promoter reaches half its strength; $\eta$ is the cooperativity in the auto-regulation promoter; $k_0$ is the promoter leakage rate; and $u(UV(t))$ is the T7 promoter strength. 
Feedback enables the external modulation of the light input using the state of the system, thereby controlling the T7 promoter strength as a function of state rather than time and eliminating the explicit time dependance in $u(UV(t))$.

Protein degradation is assumed to be a first order process with rate $\gamma$:
\begin{equation}
\nu_2(x) = \gamma x.
\end{equation}
Parameters describing the auto-regulation promoter, $\kappa$, $\eta$, $\beta$, and $k_0$, and the degradation rate $\gamma$
%were chosen to reproduce dynamics measured in \cite{Baumschlager2017} and 
% XXX - I don't think that this is right.  Only the parameters u(UV) and gamma were fit to data.
% MMM - yes that is right
are presented in Table 1~\cite{May2021}.

@@

The stoichiometry and propensities allow us to use the Gillespie stochastic simulation algorithm\cite{Gillespie1992, Gillespie1977} to generate representative trajectories of the stochastic process. However, a direct analysis of the chemical master equation (CME) is necessary for performance quantification and optimal controller design. The CME, based on the same stoichiometry and propensity functions, is a linear ODE that describes the time-dependent changes in probability mass due to the specified reaction propensities and stoichiometries. It can be expressed as:
\begin{equation}
\frac{d}{dt}P(\mathbf{X}_i)=\sum_{\mu=1}^{2{\rm N_c}}
\left(
-w_{\mu}(\mathbf{X}_i)P(\mathbf{X}_i)
+w_{\mu}(\mathbf{X}_i-\mathbf{s}_{\mu})P(\mathbf{X}_i-\mathbf{s}_{\mu})
\right).\label{CMEindex}
\end{equation}
For convenience, the CME can also be formulated more compactly in matrix/vector form as:
\begin{equation}
\frac{d}{dt}\mathbf{P}=(\mathbf{A}_0+\textbf{Bu}^{\mathcal{C}})\mathbf{P},\label{CME}
\end{equation}
where $\mathbf{P} = [P(\mathbf{X}_1), P(\mathbf{X}_2), \ldots ]^T$ is the enumerated probability mass vector for all possible states of the system; $\mathbf{A}_0$ is the infinitesimal generator of the stochastic process due to the autoregulation promoter and degradation events;  $\textbf{u}^{\mathcal{C}} =[u^{\mathcal{C}}(\mathbf{X}_1), u^{\mathcal{C}}(\mathbf{X}_2), \ldots ]^T$ is the collection of control inputs associated with each state; and $\textbf{Bu}^{\mathcal{C}}$ is the contribution that these control inputs make to the infinitesimal generator when included into the feedback process.

@@

In the SIMO control of the multiple cell process, any input signal applied to one cell affects all cells in the population. Thus, it is not possible to independently control individual cells without impacting others. An effective controller must strike a balance among the desired behaviors of all cells in the system.

@@

We studied two controller designs: the fully aware controller ($\mathbf{u}^{FAC}$) that utilizes simultaneous protein count observations from both cells, and the partially aware controller ($\mathbf{u}^{PAC}$) that relies on observations from a single cell.
\begin{equation}
\mathbf{u}^{FAC}=u(x_1,x_2)
\end{equation}
\begin{equation}
\mathbf{u}^{PAC}=u(x_2)
\end{equation}
% XXX - Do we need to use 0 and 1.  Why not use 1 and 2 so that this can match better to the formulaton above?  I changed them to x1 and x2 here, but please chekc to make sure that this is sonsistent throughout text and figures.
% MMM - Made the change to x1 x2
where $x_1$ and $x_2$ are discrete integers greater than or equal to zero that represent the number of proteins in cell one and cell two. In \cite{May2021}, we optimized the controllers $\mathbf{u}^{FAC}$ and $\mathbf{u}^{PAC}$ to minimize $J$ and saved them as lookup tables.

\subsection{Scaling for system granularity}
Previous studies have addressed the computational challenge of analyzing discrete stochastic chemical kinetics by projecting the FSP onto lower-dimensional spaces using techniques such as time scale separations \cite{Peles2006}, Krylov subspaces \cite{Macnemara&Sidje?}, coarse meshes \cite{Munsky2008IEEE,Tapia2012CDC}, or principle orthogonal decompositions \cite{Vo2019}. However, using CME analyses is often hindered by an experimental limitations.  Gene expression quantification based on fluorescence intensities, such as using fluorescent protein reporters, typically captures relative changes in expression over time rather than providing direct molecular counts.
% XXX - would be good to look for a few other examples outside of our own papers where the CME dimension is reduced by linear projections.

It would be convenient to create a parameter that could increase or decrease the intrinsic noise of a system without changing the overall dynamics of the system. This would enable the analysis of models with more or less noise without the need of new unique models each time.  A rescaling of the system could decrease the intrinsic noise while maintaining the same general behavior. 
To accomplish this rescaling, let $M$ denote the size of the origional unscaled system and let $M'$ denote the size of the newly scaled system.

We define a granularity parameter ($\alpha=M'/M$) that linearly scales the species populations to increase ($\alpha>1$) or decrease ($\alpha<1$) the maximum populations size, while maintaining the dynamics and general behavior of the model. To apply the granularity parameter, we assume that each propensity function, $w_{\mu}$, from Eqns.\ \ref{Props} is rescaled to a different level of discreteness by substituting
\begin{equation}
w'(\mathbf{X})=w(\mathbf{X}/\alpha).
\end{equation}
For example, the production and degradation of protein in cell one would become:
\begin{equation}
w'_1(\mathbf{X}) = u(\mathbf{X}/\alpha,t)  + \kappa \frac{(x_1/\alpha)^\eta}{(x_1/\alpha)^{\eta}+\beta^{\eta}} + k_0,\ w'_2(\mathbf{X}) = \gamma x_1/\alpha.
\end{equation}
% XXX - Please check this against what you did computationally.

We note that in transforming the propensity functions, the inputs to the controller must also be scaled by $1/\alpha$ before computing the appropriate functions of that state, e.g., ${u}^{\mathcal{C}} = {u}^{\mathcal{C}}(\mathbf{X}/\alpha)$. Because identification of the original control formulation, $\mathbf{u}^{FAC}(x_1,x_2)$ and $\mathbf{u}^{PAC}(x_1)$ only considered integer values for $(x_1,x_2)$, fractional number inputs for after rescaling $(x_1/\alpha,x_2/\alpha)$ are handled by 2D cubic interpolation of the nearest values.
%\begin{equation}
%u'=u(x_0 / \alpha, x_1 / \alpha) 
%\end{equation}
%\begin{equation}
%u'=u(\alpha x_0)
%\end{equation}
Finally, to provide a consistency relative scoring, the definition of the performance score is also adjusted according to sale magnitudes, i.e.,:
{\begin{align}
J' &= \sum_{i=0}^{M'}  \sum_{j=0}^{M'}P^*(x_1=i,x_2=j) ((i/\alpha - \mathbf T_1)^2 + (j/\alpha -\mathbf T_2)^2),\nonumber \\
& =\mathbf{C}'\mathbf{P}^*
\label{EuclidV}
\end{align}}

\subsection{Observation and actuation time delays}

Delays are inherent in optogenetic control systems due to biochemical reactions involved in observations, data analysis, decision making, and actuation dynamics. To investigate the effects of time delay on control actuation, we devised a time-delay stochastic simulation algorithm. This algorithm records the state history after each stochastic jump, enabling reconstruction of the population history of the species and control input.  Using this information, the time-delayed control signal at time $t$ can be specified as:

\begin{equation}
u_{\tau}(t)=\left\{
\begin{array}{rl}
      0 ,&\text{ for }  t \leq \tau, \\
      {u}^{\mathcal{C}}(x_1(t-\tau), x_2(t-\tau)) , &\text{ for }   t > \tau,\\
\end{array}\right. 
\label{timeDelaySSA}
\end{equation}
where $\tau$ is the time delay between observation and actuation, and ${u}^{\mathcal{C}}$ is the previously optimized control law (e.g., $\mathbf{u}^{\rm FAC}$ or $\mathbf{u}^{\rm PAC}$). We note that the time delay stochastic process was only simulated using the SSA because to our knowledge an appropriate direct FSP/CME integration procedure has not yet been developed.

@@

\subsection{Dynamic setpoint control}

static setpoint control is the control of a reference point which does not move over time. 
dynamic setpont control targets a mvonig referenc peoint.
synamis serpoint control is hard to deriuve, but the optimization of peicesice constangtof each static referenc pont is easy and can be stiched together.

@@

\subsection{Effects of Parameter Errors or Extrinsic Uncertainties}
\begin{figure*}
\begin{center}
\includegraphics[width=1\textwidth]{ParameterPerturbation.pdf}
\caption{Parameter sweeps using the FAC and PAC show a broad range of control performance in cell 1 ({\bf A}),  in cell 2 ({\bf B}), and in both cells ({\bf C}). Columns show each parameter in the model, rows show the cell which has its parameter perturbed. }
\label{Parameter}
\end{center}
\end{figure*}
